### 2023-12-31 sections/2-methods.tex
fix reference
squashed all red TODOs

In this manuscript, we are primarily concerned with $k$-NN search in a finite-dimensional space.
Given a dataset $\textbf{X} = \{x_1 \dots x_n\}$ of cardinality $|\textbf{X}| = n$, we define a \emph{point} or \emph{datum} $x_i \in \textbf{X}$ as a singular observation, for example, the neural-network embedding of an image, the genome of an organism, a measurement of a radio frequency signal, etc.

We define a \emph{distance function} $f : \textbf{X} \times \textbf{X} \mapsto \mathbb{R}^+ \ \cup \ \{0\}$ which, given two points, deterministically returns a finite and non-negative real number.
A distance value of zero defines an identity among points (i.e. $f(x, y) = 0 \Leftrightarrow x = y$) and larger values indicate greater dissimilarity among the points.
We also require that the distance function be symmetric (i.e., $f(x, y) = f(y, x) \ \forall \ x, y \in \textbf{X}$).
In addition to these constraints, if the distance function obeys the triangle inequality (i.e. $f(x, y) \leq f(x, z) + f(z, y) \ \forall \ x, y, z \in \textbf{X}$), then it is also a \emph{distance metric}.
Under distance metrics, all search algorithms in CAKES are exact.
For example, Euclidean, Levenshtein~\cite{levenshtein1966binary} and Dynamic Time Warping (DTW)~\cite{muller2007dynamic} distances are all distance metrics while Cosine distance is not a metric because it does not obey the triangle inequality (e.g. consider the points $x = (1, 0)$, $y = (0, 1)$ and $z = (1, 1)$ on a Cartesian plane).

@@

CAKES assumes the manifold hypothesis~\cite{fefferman2016testing}, i.e., high-dimensional data collected from constrained generating phenomena typically only occupy a low-dimensional manifold within their embedding space.
We say that such data are \emph{manifold-constrained}, and have low \emph{local fractal dimension} (LFD).
In other words, we assume that the dataset is embedded in a $D$-dimensional space, but that the data only occupy a $d$-dimensional manifold, where $d \ll D$.
While we sometimes use Euclidean notions, such as voids, volumes and proximity, to talk about the geometric and topological properties of the clusters and manifold, CLAM and CAKES do not rely on such notions;
they serve merely as convenient and intuitive vocabulary to discuss the underlying mathematics.

@@

where $B_X(q, r)$ is the set of points contained in the metric ball of radius $r$ centered at a point $q$ in the dataset $\textbf{X}$.
We stress that this concept differs from the \emph{embedding dimension} of a dataset.
To illustrate the difference, consider the Silva 18S rRNA dataset which contains which contains genomes with unaligned lengths of up to 3,718 base pairs.
Hence, the \emph{embedding dimension} of this dataset is at least 3,718.
However, due to physical constraints (namely, the laws that govern RNA replication and mutation), the data are constrained to a lower-dimensional manifold within this embedding space.
LFD is an approximation of the dimensionality of that lower-dimensional manifold in the ``vicinity'' of a given point.
Figure~\ref{fig:results:lfd-plots} illustrates this concept on a variety of datasets, showing how real datasets obey the manifold hypothesis.

@@

We define a \emph{cluster} as a set of points with a \emph{center} and a \emph{radius}.
The \emph{center} is the geometric median of the points (or a smaller sample of the points) in the \emph{cluster}, and so it is a real data point.
The \emph{radius} is the maximum distance from the \emph{center} to any point in the \emph{cluster}.
Each non-leaf cluster has two child clusters in much the same way that a node in a binary tree has two child nodes.
Hereafter, when we refer to LFD, it is estimated at the scale of the cluster radius and half that radius.
This allows us to rewrite Definition~\ref{eq:methods:lfd-original} as:

\begin{equation}
    \log_2 \left( \frac{|B_X(q, r)|}{|B_X(q, r/2)|} \right)
    \label{eq:methods:lfd-simplified}
\end{equation}

The \emph{metric entropy} ($\mathcal{N}_{r}(X)$) for some radius $r$ was defined, in~\cite{yu2015entropy}, as the minimum number of clusters of a uniform radius $r$ needed to cover the data.
In this paper, we define the metric entropy $\mathcal{N}_{\hat{r}}(X)$ of a dataset $X$ for the hierarchical clustering scheme as the number of leaf clusters in the tree where $\hat{r}$ is the mean radius of all leaf clusters.

@@

For a cluster $C$ with $|C|$ points, we begin by taking a random sample of $\sqrt{|C|}$ of its points, and computing pairwise distances between all points in this sample.
Using these distances, we compute the \emph{geometric median} of this sample; in other words, we find the point which minimizes the sum of distances to all other points in the sample.
This geometric median is the \emph{center} of $C$.

The \emph{radius} of $C$ is the maximum distance from the \emph{center} to any other point in $C$.
The point which is responsible for that radius (i.e., the furthest point from the center) is designated the \emph{left pole} and the point which is furthest from left pole is designated the \emph{right pole}.

We then partition the cluster into a \emph{left child} and a \emph{right child}, where the left child contains all points in the cluster which are closer to the left pole than to the right pole, and the right child contains all points in the cluster which are closer to the right pole than to the left pole.
Without loss of generality, we assign to the left child those points which are equidistant from the two poles.

@@

        \STATE $seeds \Leftarrow \lfloor \sqrt{|C|} \rfloor$ random points from $C$
        \STATE $c \Leftarrow$ geometric median of $seeds$
        \STATE $l \Leftarrow \argmax f(c, x) \ \forall \ x \in C$
        \STATE $r \Leftarrow \argmax f(l, x) \ \forall \ x \in C$
        \STATE $L \Leftarrow \{x \ | \ x \in C \land f(l, x) \le f(r, x)\}$
        \STATE $R \Leftarrow \{x \ | \ x \in C \land f(r, x) < f(l, x)\}$

@@

In CHESS, each cluster stored a list of indices into the dataset.
This list was used to retrieve the clusters' points during search.
Although this approach allowed us to retrieve the points in constant time, its memory cost was prohibitively high.
With a dataset of cardinality $n$ and each cluster storing a list of indices for its points, we stored a total of $n$ indices at each depth in the tree.
Assuming a balanced tree, and thus $\mathcal{O}(\log(n))$ depth, this approach had a memory overhead of $\mathcal{O}(n\log(n))$.

One may improve this memory overhead to $\mathcal{O}(n)$ by only storing indices at the leaf clusters.
This approach, however, introduces a time cost of $\mathcal{O}(n\log(n))$ whenever we need to find the indices for a non-leaf cluster (because it requires a traversal of the subtree rooted at that cluster).

In this work, we introduce a new approach wherein, after building the cluster tree, we reorder the dataset so that points are stored in a depth-first order.
Then, within each cluster, we need only store its \emph{cardinality} and an \emph{offset} to access the its points from the dataset.
The root cluster has an offset of zero and a cardinality equal to the number of points in the dataset.
A left child has the same offset as that of its parent, and the corresponding right child has an offset equal to the left child's offset plus the left child's cardinality.

With no additional memory cost nor time cost for retrieving points during search, depth-first reordering offers significant improvements over the approach used in CHESS.

@@

% While it may be tempting to assume that cluster radii decrease with each application of Partition (refer to Algorithm \ref{alg:methods:partition}), unfortunately, this assumption is incorrect. 
% Fortunately, we \emph{can} make some guarantees about the scaling behavior of cluster radii;
% in particular, we prove in this section that cluster radii will stop increasing after at most $d$ partitions, where $d$ is the fractal dimensionality of the dataset.

% We can describe a $d$-dimensional distribution of data by choosing some set of $d$ mutually orthogonal axes.
% Let $2R$ be the maximum pairwise distance among the points in the dataset. 
% We choose the axes such that the two points that are $2R$ apart lie \emph{along one of the axes}. 
% Thus, a $d$-dimensional hyper-sphere of radius $R$ would bound the dataset. 
% In the worst case, (i.e., with a uniform-density distribution that fills the $d$-sphere), our axes will be such that $2R$ is the maximum pairwise distance \emph{along every axis}. 
% Such a distribution would also produce a balanced cluster tree.

@@

Given a query $q$, along with a distance function $f$ defined on a dataset $\textbf{X}$, $k$-NN search aims to find the set $H$ such that $|H| = k$ and $\forall x \in \textbf{X} \setminus H, \ f(x, q) > \max \{ f(y, q): y \in H \}$;
in other words, $k$-NN finds the $k$ closest points to $q$ in $\textbf{X}$.
We also have the $\rho$-NN search problem, which aims to find the set $\{x \in \textbf{X}: f(q, x) \leq \rho \}$;
that is, find all points in $\textbf{X}$ that are no more than a distance $\rho$ from $q$.

Given a Cluster $C$, let $c$ be its center and $r$ be its radius.
Our $\rho$-NN and $k$-NN search algorithms make use of the following properties, as illustrated in Figure~\ref{fig:methods:deltas}.

\begin{itemize}
    \item $\delta = f(q, c)$ is the distance from the query to the cluster center $c$.
    \item $\delta^{+} = \delta + r$ is the distance from the query to the theoretically farthest point in $C$.
    \item $\delta^{-} = \text{max}(0, \delta - r)$ is the distance from the query to the theoretically closest point in $C$.
\end{itemize}

@@

We define a \emph{singleton} as a cluster which either contains a single point (i.e., has cardinality 1) or which contains duplicates of the same point (i.e., has cardinality greater than 1 but contains only one \emph{unique} point).
A singleton clearly has zero radius, and so $\delta = \delta^{-} = \delta^{+}$.
Hence, we overload the above notation to also refer to the distance from a query to an individual point.

@@

Let $q$ denote the query, $\rho$ denote the search radius, and $l$ and $r$ denote the cluster's left and right poles respectively (see Section~\ref{sec:methods:clustering:building-the-tree}).
Without loss of generality, we assume that $\overline{q r \vphantom{l}} \leq \overline{q l}$.
Now let $q'$ be the projection of $q$ to $\overline{l r}$, $m$ be the midpoint of $\overline{l r}$ and $d$ be the distance from $q'$ to $m$.
As a consequence of how we assign a point in the parent cluster to the left child in Algorithm~\ref{alg:methods:partition}, if $q'$ is too far from $m$, i.e. $\rho < d$, then when the left child cannot contain points inside the query ball.
In such a case we proceed only with the right child.
Otherwise, we proceed with both children.

@@

To perform $\rho$-NN search, we first perform a coarse \emph{tree-search}, as outlined in Algorithm~\ref{alg:methods:rnn-search:tree-search}, to find the leaf clusters which overlap with the query ball or any clusters which lie entirely within the query ball.
Then, for all such clusters, we perform a finer-grained \emph{leaf-search}, as outlined in Algorithm~\ref{alg:methods:rnn-search:leaf-search}, to find all points in which are no more than a distance $\rho$ from the query.

@@

            \STATE $[l, r]$ $\Leftarrow$ \emph{pole}s of $C$
            \STATE $[L, R]$ $\Leftarrow$ \emph{children} of $C$

@@

\begin{gather}
    \mathcal{O}
    \Bigg(
        \underbrace{\log_2 \mathcal{N}_{\hat{r}}(X)}_{\textrm{metric entropy}} +
        \overbrace{ \left| B_X(q, \rho) \right|}^{\textrm{output size}}
        \underbrace{ \left( 1 + \frac{2 \cdot \hat{r}}{ \rho} \right) ^ d}_{\textrm{scaling factor}}
    \Bigg)
    \label{eq:methods:rnn-search-complexity}
\end{gather}

Here, $\hat{r}$ is the \textit{mean} radius of leaf clusters, $\mathcal{N}_{\hat{r}}(X)$ is the metric entropy at that radius, $B_X(q, \rho)$ is a ball of radius $\rho$ around the query $q$, and $d$ is the LFD around the query in a region of radius $\rho + \hat{r}$.
From Algorithm~\ref{alg:methods:rnn-search}, we have that $H = B_X(q, \rho)$, and so $\rho$-NN search performance scales linearly with the size of the output set and exponentially with the LFD around the query.
While one might worry that the exponential scaling factor would dominate, the manifold hypothesis suggests that the LFD of real-life datasets is typically very low, and so this algorithm actually scales sub-linearly with the cardinality of the dataset.

@@

In these algorithms, we use $H$, for \emph{hits}, to refer to the data structure which stores the closest points to the query found so far, and $Q$ to refer to the data structure which stores the clusters and points which are still in contention for being one of the $k$ nearest neighbors.

@@

In this algorithm, we perform $\rho$-nearest-neighbors search starting with a small search radius, and repeatedly increasing the radius until $|H| \geq k$.

Let the search radius, $r$, be equal to the radius of the root cluster divided by the cardinality of the dataset.
We perform tree-search with radius $r$.
If no clusters are returned, then we double $r$ and perform tree-search again, repeating until we find at least one cluster.
Let $S$ be the set of clusters returned by the first tree-search which returns at least one cluster.

Now, so long as $\sum_{C \in S} |C| < k$, we continue to perform tree-search, but instead of doubling $r$ on each iteration, we multiply it by a factor determined by the LFD in the vicinity of the query ball. 
In particular, we increase the radius by a factor of

\begin{equation}
    \min \left(2, \left( {\frac{k}{\sum_{C \in S} |C|}} \right)^{\mu^{-1}} \right)
    \label{eq:methods:repeated-rnn-factor}
\end{equation}

@@

Intuitively, the factor by which we increase the radius should be \emph{inversely} related to the number of points found so far. 
Additionally, when the LFD at the radius scale from the previous iteration is low, this suggests that the data are relatively concentrated in that region.
In such a region, a small increase in the radius would likely encounter vacant space, so a larger radial increase is needed.
Thus, the factor of radius increase should also be \emph{inversely} related to the LFD.

Once $\sum_{C \in S} |C| \geq k$, we are guaranteed to have found at least $k$ neighbors, and so we can stop increasing the radius.
We perform $\rho$-NN search with this radius and return the $k$ nearest neighbors.

\begin{algorithm} % enter the algorithm environment
    \caption{repeated-$\rho$-NN($root$, $q$, $k$)} % give the algorithm a caption
    \label{alg:methods:repeated-rnn} % and a label for \ref{} commands later in the document
    \begin{algorithmic} % enter the algorithmic environment
        \REQUIRE $root$, the root cluster
        \REQUIRE $q$, a query
        \REQUIRE $k$, the number of neighbors to find

        \STATE $r \Leftarrow radius$ of the $root$ cluster
        \STATE $r \Leftarrow$ $\frac{r}{|root|}$
        \STATE $Q \Leftarrow \emptyset$, a set of clusters

        \WHILE {$Q = \emptyset$}
            \STATE $Q \Leftarrow$ tree-search($root$, $q$, $r$)
            \STATE $r \Leftarrow 2 \cdot r$
        \ENDWHILE

@@

We find it useful to adopt the terminology used in \cite{ishaq2019clustered} and \cite{yu2015entropy}, and address \emph{tree-search} and \emph{leaf-search} separately.
Tree-search refers to the process of identifying clusters which have overlap with the query ball, or in other words, clusters which might contain one of the $k$ nearest neighbors. 
Leaf-search refers to the process of identifying the $k$ nearest neighbors among the points in the clusters identified by tree-search.

In ~\cite{ishaq2019clustered}, we showed that the complexity of tree-search is $\mathcal{O}(\log\mathcal{N}_{\hat{r}}(X))$, where $\mathcal{N}_{\hat{r}}(X)$ is the metric entropy of the dataset $X$ at a radius $\hat{r}$.
To adjust this bound for Repeated $\rho$-NN, we must estimate the number of iterations of tree-search needed to find a radius that guarantees at least $k$ neighbors.

Based on the assumption that the LFD near the query does not differ significantly from that of nearby clusters, Equation~\ref{eq:methods:repeated-rnn-factor} suggests that in the expected case, we need only two iterations of tree-search to find $k$ neighbors:
one iteration to find at least one neighbor, and the one more to find enough remaining neighbors.
Since this is a constant factor, complexity of tree search for Repeated $\rho$-NN is the same as that of $\rho$-NN.

To determine the asymptotic complexity of leaf-search, we must estimate $\sum_{C \in Q} |C|$, the number of points in the clusters returned by tree search.
Since we must examine each point in each such cluster, time complexity of leaf-search is linear in $\sum_{C \in Q} |C|$. 
Let $\rho_k$ be the distance from the query to the $k$th nearest neighbor.
Then, we see that $Q$ is expected to be the set of clusters which overlap with a ball of radius $\rho_k$ around the query.
We can estimate this region as a ball of radius $\rho_k + 2\hat{r}$, where $\hat{r}$ is the mean radius of the clusters in $Q$.

@@

\begin{equation*}
    \sum_{C \in S} |C| \leq \gamma  \left| B(q, \rho_k) \right| \left(1+ \frac{2\hat{r}}{\rho_k} \right)^d
\end{equation*}

@@

\begin{equation*}
    \sum_{C \in S} |C| \leq \gamma k \left( 1+ \frac{2\hat{r}}{\rho_k} \right)^d
\end{equation*}

@@

We let $\hat{d}$ be the mean LFD of the clusters in $Q$.
While ordinarily we compute LFD by comparing cardinalities of two balls with two different radii centered at \emph{the same} point, in order to estimate $\rho_k$, we instead compare the cardinality of a ball \emph{around the query} of radius $\rho_k$ to the mean cardinality, $\hat{|C|}$, of clusters in $Q$ at a radius equal to the mean of their radii, $\hat{r}$.

We justify this approach by noting that, since the query is from the same distribution as the rest of the data, we could move from the query to the center of one of the nearby clusters without significantly changing our estimate of the LFD.

>>>>

In this manuscript, we are primarily concerned with $k$-NN search in a finite-dimensional space.
Given a dataset $\textbf{X} = \{x_1 \dots x_n\}$ of cardinality $|\textbf{X}| = n$, we define a \textit{point} or \textit{datum} $x_i \in \textbf{X}$ as a singular observation, for example, the neural-network embedding of an image, the genome of an organism, a measurement of a radio frequency signal, etc.

We define a \textit{distance function} $f : \textbf{X} \times \textbf{X} \mapsto \mathbb{R}^+ \ \cup \ \{0\}$ which, given two points, deterministically returns a finite and non-negative real number.
A distance value of zero defines an identity among points (i.e. $f(x, y) = 0 \Leftrightarrow x = y$) and larger values indicate greater dissimilarity among the points.
We also require that the distance function be symmetric (i.e., $f(x, y) = f(y, x) \ \forall \ x, y \in \textbf{X}$).
In addition to these constraints, if the distance function obeys the triangle inequality (i.e. $f(x, y) \leq f(x, z) + f(z, y) \ \forall \ x, y, z \in \textbf{X}$), then it is also a \textit{distance metric}.
Under distance metrics, all search algorithms in CAKES are exact.
For example, Euclidean, Levenshtein~\cite{levenshtein1966binary} and Dynamic Time Warping (DTW)~\cite{muller2007dynamic} distances are all distance metrics while Cosine distance is not a metric because it does not obey the triangle inequality (e.g. consider the points $x = (1, 0)$, $y = (0, 1)$ and $z = (1, 1)$ on the Cartesian plane).

@@

CAKES assumes the manifold hypothesis~\cite{fefferman2016testing}, i.e., high-dimensional data collected from constrained generating phenomena typically only occupy a low-dimensional manifold within their embedding space.
We say that such data are \textit{manifold-constrained} and have low \textit{local fractal dimension} (LFD).
In other words, we assume that the dataset is embedded in a $D$-dimensional space, but that the data only occupy a $d$-dimensional manifold, where $d \ll D$.
While we sometimes use Euclidean notions, such as voids, volumes and proximity to describe the geometric and topological properties of the clusters and manifold, CLAM and CAKES do not rely on such notions;
they serve merely as convenient and intuitive vocabulary to discuss the underlying mathematics.

@@

where $B_X(q, r)$ is the set of points contained in the metric ball of radius $r$ centered at a point $q$ in the dataset $\textbf{X}$.
Intuitively, LFD measures the rate of change in the number of points in a ball of radius $r$ around a point $q$ as $r$ increases.

We stress that this concept differs from the \textit{embedding dimension} of a dataset.
To illustrate the difference, consider the Silva 18S rRNA dataset which contains which contains genomes with unaligned lengths of up to 3,718 base pairs and aligned length of 50,000 base pairs.
Hence, the \textit{embedding dimension} of this dataset is at least 3,718 and at most 50,000.
However, due to physical constraints (namely, biological evolution and the chemistry of RNA replication and mutation processes), the data are constrained to a lower-dimensional manifold within this embedding space.
LFD is an approximation of the dimensionality of that lower-dimensional manifold in the ``vicinity'' of a given point.
Figure~\ref{fig:results:lfd-plots} illustrates this concept on a variety of datasets, showing how real datasets obey the manifold hypothesis.

@@

We define a \textit{cluster} as a set of points with a \textit{center} and a \textit{radius}.
The \textit{center} is the geometric median of the points (or a smaller sample of the points) in the \textit{cluster}, and so it is a real data point.
The \textit{radius} is the maximum distance from the \textit{center} to any point in the \textit{cluster}.
Each non-leaf cluster has two child clusters in much the same way that a node in a binary tree has two child nodes.
Note that clusters can have overlapping volumes and, in such cases, points in the overlapping volume are assigned to exactly one of the overlapping clusters.
As a consequence, a cluster can be a subset of the metric ball at the same center and radius, i.e. $C(c, r) \subseteq B_X(c, r)$.

Hereafter, when we refer to the LFD of a \textit{cluster}, it is estimated at the length scale of the cluster radius and half that radius.
This allows us to rewrite Definition~\ref{eq:methods:lfd-original} as:

\begin{equation}
    \log_2 \left( \frac{|C(c, r)|}{|C(c, \frac{r}{2})|} \right)
    \label{eq:methods:lfd-simplified}
\end{equation}

where $|C(c, r)|$ is the cardinality of the cluster $C$ with center $c$ and radius $r$, and $C(c, \frac{r}{2})$ is the set of points in $C$ which are no more than a distance $\frac{r}{2}$ from $c$, i.e. $C(c, \frac{r}{2}) = \{p : p \in C \land f(c, p) \leq \frac{r}{2}\}$.

The \textit{metric entropy} $\mathcal{N}_{r}(X)$ for some radius $r$ was defined, in~\cite{yu2015entropy}, as the minimum number of clusters of a uniform radius $r$ needed to cover the data.
In this paper, we define the metric entropy $\mathcal{N}_{\hat{r}}(X)$ of a dataset $X$ for the hierarchical clustering scheme as the number of leaf clusters in the tree where $\hat{r}$ is the mean radius of all leaf clusters.

@@

For a cluster $C$ with $|C|$ points, we begin by taking a random sample of $\sqrt{|C|}$ of its points, and computing pairwise distances between all points in this sample.
Using these distances, we compute the \textit{geometric median} of this sample; in other words, we find the point which minimizes the sum of distances to all other points in the sample.
This geometric median is the \textit{center} of $C$.

The \textit{radius} of $C$ is the maximum distance from the \textit{center} to any other point in $C$.
The point which is responsible for that radius (i.e., the furthest point from the center) is designated the \textit{left pole} and the point which is furthest from left pole is designated the \textit{right pole}.

We then partition the cluster into a \textit{left child} and a \textit{right child}, where the left child contains all points in the cluster which are closer to the left pole than to the right pole, and the right child contains all points in the cluster which are closer to the right pole than to the left pole.
Without loss of generality, we assign to the left child those points which are equidistant from the two poles.

@@

        \STATE $seeds \Leftarrow$ random sample of $\left\lceil \sqrt{|C|} \right\rceil$ points from $C$
        \STATE $c \Leftarrow$ geometric median of $seeds$
        \STATE $l \Leftarrow \argmax f(c, x) \ \forall \ x \in C$
        \STATE $r \Leftarrow \argmax f(l, x) \ \forall \ x \in C$
        \STATE $L \Leftarrow \{x \ | \ x \in C \land f(l, x) \le f(r, x)\}$
        \STATE $R \Leftarrow \{x \ | \ x \in C \land f(r, x) < f(l, x)\}$

@@

In CHESS, each cluster stored a list of indices into the dataset.
This list was used to retrieve the clusters' points during search.
Although this approach allowed us to retrieve the points in constant time, its memory cost was prohibitively high.
With a dataset of cardinality $n$ and each cluster storing a list of indices for its points, we stored a total of $n$ indices at each depth in the tree.
Assuming a balanced tree, and thus $\mathcal{O}(\log n)$ depth, this approach had a memory overhead of $\mathcal{O}(n \log n)$.

One may improve this memory overhead to $\mathcal{O}(n)$ by only storing indices at the leaf clusters.
This approach, however, introduces a time cost of $\mathcal{O}(n \log n)$ whenever we need to find the indices for a non-leaf cluster (because it requires a traversal of the subtree rooted at that cluster).

In this work, we introduce a new approach wherein, after building the cluster tree, we reorder the dataset so that points are stored in a depth-first order.
Then, within each cluster, we need only store its \textit{cardinality} and an \textit{offset} to access the its points from the dataset.
The root cluster has an offset of zero and a cardinality equal to the number of points in the dataset.
A left child has the same offset as that of its parent, and the corresponding right child has an offset equal to the left child's offset plus the left child's cardinality.

With no additional memory cost nor time cost for retrieving points during search, depth-first reordering offers significant improvements at search-time over the approach used in CHESS.

@@

% While it may be tempting to assume that cluster radii decrease with each application of Partition (refer to Algorithm \ref{alg:methods:partition}), unfortunately, this assumption is incorrect. 
% Fortunately, we \textit{can} make some guarantees about the scaling behavior of cluster radii;
% in particular, we prove in this section that cluster radii will stop increasing after at most $d$ partitions, where $d$ is the fractal dimensionality of the dataset.

% We can describe a $d$-dimensional distribution of data by choosing some set of $d$ mutually orthogonal axes.
% Let $2R$ be the maximum pairwise distance among the points in the dataset. 
% We choose the axes such that the two points that are $2R$ apart lie \textit{along one of the axes}. 
% Thus, a $d$-dimensional hyper-sphere of radius $R$ would bound the dataset. 
% In the worst case, (i.e., with a uniform-density distribution that fills the $d$-sphere), our axes will be such that $2R$ is the maximum pairwise distance \textit{along every axis}. 
% Such a distribution would also produce a balanced cluster tree.

@@

Given a query $q$, along with a distance function $f$ defined on a dataset $\textbf{X}$, we define the $k$-NN search problem which aims to find the $k$ closest points to $q$ in $\textbf{X}$;
i.e. the set $H$ such that $|H| = k$ and $H = B_X(q, \rho_k)$ where $\rho_k = \max \left\{ f(q, p) \ \forall \ p \in H \right\}$ is the distance from $q$ to the $k^{th}$ nearest neighbor in $\textbf{X}$.
We also have the $\rho$-NN search problem, which aims to find the all points in $\textbf{X}$ that are no more than a distance $\rho$ from $q$;
i.e. the set $H = B_X(q, \rho) = \{x \in \textbf{X}: f(q, x) \leq \rho \}$.

Given a cluster $C$, let $c$ be its center and $r$ be its radius.
Our $\rho$-NN and $k$-NN search algorithms make use of the following properties, as illustrated in Figure~\ref{fig:methods:deltas}.

\begin{itemize}
    \item $\delta = f(q, c)$, the distance from the query to the cluster center $c$.
    \item $\delta^{+} = \delta + r$, the distance from the query to the theoretically farthest point in $C$.
    \item $\delta^{-} = \text{max}(0, \delta - r)$, the distance from the query to the theoretically closest point in $C$.
\end{itemize}

@@

We define a \textit{singleton} as a cluster which either contains a single point (i.e., has cardinality 1) or which contains duplicates of the same point (i.e., has cardinality greater than 1 but contains only one \textit{unique} point).
A singleton clearly has zero radius, and so $\delta = \delta^{-} = \delta^{+}$.
Hence, we overload the above notation to also refer to the distance from a query to an individual point.

@@

Let $q$ denote the query, $\rho$ denote the search radius, and $l$ and $r$ denote the cluster's left and right poles respectively (see Section~\ref{sec:methods:clustering:building-the-tree}).
Without loss of generality, we assume that $\overline{q r \vphantom{l}} \leq \overline{q l}$.
Now let $q'$ be the projection of $q$ to $\overline{l r}$, $m$ be the midpoint of $\overline{l r}$, and $d$ be the distance from $q'$ to $m$.
As a consequence of how we assign a point in the parent cluster to the left child in Algorithm~\ref{alg:methods:partition}, if $q'$ is too far from $m$, i.e. $\rho < d$, then when the left child cannot contain points inside the query ball.
In such a case we proceed only with the right child.
Otherwise, we proceed with both children.

@@

To perform $\rho$-NN search, we first perform a coarse \textit{tree-search}, as outlined in Algorithm~\ref{alg:methods:rnn-search:tree-search}, to find the leaf clusters which overlap with the query ball or any clusters which lie entirely within the query ball.
Then, for all such clusters, we perform a finer-grained \textit{leaf-search}, as outlined in Algorithm~\ref{alg:methods:rnn-search:leaf-search}, to find all points in which are no more than a distance $\rho$ from the query.

@@

            \STATE $[l, r]$ $\Leftarrow$ \textit{pole}s of $C$
            \STATE $[L, R]$ $\Leftarrow$ \textit{children} of $C$

@@

\begin{gather}
    \mathcal{O}
    \Bigg(
        \underbrace{
            \log~\overbrace{\mathcal{N}_{\hat{r}}(X)}^{\textrm{metric entropy}}
        }_{\textrm{tree search}}
        \ + \ 
        \underbrace{
            \overbrace{ \big| B_X(q, \rho) \big|}^{\textrm{output size}}
            \overbrace{ \left( \frac{\rho + 2 \cdot \hat{r}}{ \rho} \right) ^ d}^{\textrm{scaling factor}}
        }_{\textrm{leaf search}}
    \Bigg)
    \label{eq:methods:rnn-search-complexity}
\end{gather}

where $\hat{r}$ is the \textit{mean} radius of leaf clusters, $\mathcal{N}_{\hat{r}}(X)$ is the metric entropy at that radius, $B_X(q, \rho)$ is a ball of radius $\rho$ around the query $q$, and $d$ is the LFD around the query at the length scale of $\rho$ and $\rho + 2 \cdot \hat{r}$.

From Algorithm~\ref{alg:methods:rnn-search}, we have that $H = B_X(q, \rho)$, and so $\rho$-NN search performance scales linearly with the size of the output set and exponentially with the LFD around the query.
While one might worry that the exponential scaling factor would dominate, the manifold hypothesis suggests that the LFD of real-life datasets is typically very low, and so this algorithm actually scales sub-linearly with the cardinality of the dataset.

@@

In these algorithms, we use $H$, for \textit{hits}, to refer to the data structure which stores the closest points to the query found so far, and $Q$ to refer to the data structure which stores the clusters and points which are still in contention for being one of the $k$ nearest neighbors.

@@

In this algorithm, we perform $\rho$-NN search starting with a small search radius, and repeatedly increasing the radius until $|H| \geq k$.

Let the search radius $r$ be equal to the radius of the root cluster divided by the cardinality of the dataset.
We perform tree-search with radius $r$.
If no clusters are found, then we double $r$ and perform tree-search again, repeating until we find at least one cluster.
Let $Q$ be the set of clusters returned by the first tree-search which returns at least one cluster.

Now, so long as $\sum_{C \in Q} |C| < k$, we continue to perform tree-search, but instead of doubling $r$ on each iteration, we multiply it by a factor determined by the LFD in the vicinity of the query ball. 
In particular, we increase the radius by a factor of

\begin{equation}
    \min \left(2, \left( {\frac{k}{\sum_{C \in Q} |C|}} \right)^{\mu^{-1}} \right)
    \label{eq:methods:repeated-rnn-factor}
\end{equation}

@@

Intuitively, the factor by which we increase the radius should be \textit{inversely} related to the number of points found so far. 
When the LFD at the radius scale from the previous iteration is high, this suggests that the data are densely populated in that region.
Thus, a small increase in the radius would likely encounter many more points, so a smaller radial increase would suffice to find $k$ neighbors.
Conversely, when the LFD at the radius scale from the previous iteration is low, this suggests that the data are sparsely populated in that region.
In such a region, a small increase in the radius would likely encounter vacant space, so a larger radial increase is needed.
Thus, the factor of radius increase should also be \textit{inversely} related to the LFD.

Once $\sum_{C \in Q} |C| \geq k$, we are guaranteed to have found at least $k$ neighbors, and so we can stop increasing the radius.
We perform $\rho$-NN search with this radius and return the $k$ nearest neighbors.

\begin{algorithm} % enter the algorithm environment
    \caption{Repeated $\rho$-NN($root$, $q$, $k$)} % give the algorithm a caption
    \label{alg:methods:repeated-rnn} % and a label for \ref{} commands later in the document
    \begin{algorithmic} % enter the algorithmic environment
        \REQUIRE $root$, the root cluster
        \REQUIRE $q$, a query
        \REQUIRE $k$, the number of neighbors to find

        \STATE $r \Leftarrow radius$ of the $root$ cluster
        \STATE $r \Leftarrow$ $\frac{r}{|root|}$

        \ALOOP{}
            \STATE $Q \Leftarrow$ tree-search($root$, $q$, $r$)
            \IF{$Q \neq \emptyset$}
                \STATE \textbf{break}
            \ENDIF
            \STATE $r \Leftarrow 2 \cdot r$
        \ENDALOOP

@@

We find it useful to adopt the terminology used in \cite{ishaq2019clustered} and \cite{yu2015entropy}, and address \textit{tree-search} and \textit{leaf-search} separately.
Tree-search refers to the process of identifying clusters which have overlap with the query ball, or in other words, clusters which might contain one of the $k$ nearest neighbors. 
Leaf-search refers to the process of identifying the $k$ nearest neighbors among the points in the clusters identified by tree-search.

In ~\cite{ishaq2019clustered}, we showed that the complexity of tree-search is $\mathcal{O}(\log\mathcal{N}_{\hat{r}}(X))$, where $\mathcal{N}_{\hat{r}}(X)$ is the metric entropy of the dataset $X$ at a radius $\hat{r}$.
To adjust this bound for Repeated $\rho$-NN, we must estimate the number of iterations of tree-search (Algorithm~\ref{alg:methods:rnn-search:tree-search}) needed to find a radius that guarantees at least $k$ neighbors.

Based on the assumption that the LFD near the query does not differ significantly from that of nearby clusters, Equation~\ref{eq:methods:repeated-rnn-factor} suggests that in the expected case, we need only two iterations of tree-search to find $k$ neighbors:
one iteration to find at least one cluster, and the one more to find enough the $k$ neighbors.
Since this is a constant factor, complexity of tree search for Repeated $\rho$-NN is the same as that of $\rho$-NN search, i.e. $\mathcal{O}\big(\log\mathcal{N}_{\hat{r}}(X)\big)$.

To determine the asymptotic complexity of leaf-search, we must estimate $\sum_{C \in Q} |C|$, the total cardinality of the clusters returned by tree search.
Since we must examine every point in each such clusters, time complexity of leaf-search is linear in $\sum_{C \in Q} |C|$. 
Let $\rho_k$ be the distance from the query to the $k^{th}$ nearest neighbor.
Then, we see that $Q$ is expected to be the set of clusters which overlap with a ball of radius $\rho_k$ around the query.
We can estimate this region as a ball of radius $\rho_k + 2\hat{r}$, where $\hat{r}$ is the mean radius of the clusters in $Q$.

@@

\begin{equation*}
    \sum_{C \in S} |C| \leq \gamma  \left| B(q, \rho_k) \right| \left(\frac{\rho_k + 2 \cdot \hat{r}}{\rho_k} \right)^d
\end{equation*}

@@

\begin{equation*}
    \sum_{C \in S} |C| \leq \gamma k \left( 1 + 2 \cdot \frac{\hat{r}}{\rho_k} \right)^d
\end{equation*}

@@

We let $\hat{d}$ be the (harmonic) mean LFD of the clusters in $Q$.
While ordinarily we compute LFD by comparing cardinalities of two balls with two different radii centered at \textit{the same} point, in order to estimate $\rho_k$, we instead compare the cardinality of a ball \textit{around the query} of radius $\rho_k$ to the mean cardinality, $\hat{|C|}$, of clusters in $Q$ at a radius equal to the mean of their radii, $\hat{r}$.
We justify this approach by noting that, since the query is from the same distribution as the rest of the data, we could move from the query to the center of one of the nearby clusters without significantly changing our estimate of the LFD.
